```{r setup, include=FALSE,message=FALSE,warning=FALSE}
# OPTIONS -----------------------------------------------
knitr::opts_chunk$set(echo = TRUE, 
                      warning=FALSE, 
                      message = FALSE)

# Tutorial packages
library(vembedr)
library(skimr)
library(yarrr)
library(RColorBrewer)
library(GGally) 
library(tidyverse)
library(plotly)
library(readxl)
library(rvest)
library(biscale)
library(tidycensus)
library(cowplot)
library(units)
```

# Reading in data {#T4_ReadingData}

<br>

## Before you start {#T4_DataCheck}

**1. Use a Project, not `setwd()`** Open your `.Rproj` file before doing anything else. R will automatically look in that project folder for your data files — no need to tell it where to look. Just make sure your data file is saved in the same folder as the `.Rproj`.

**2. Don't use File \> Import** Always use code to read in data (not the menu). The menu won't work when you knit your document.

<br>

------------------------------------------------------------------------

## Reading in your data {#T4_Commands}

Find your file type below and copy the command into a code chunk.

<br>

### Spreadsheets and tables {#T4_ReadinSpreadsheets}

<br>

#### CSV files (.csv) — `read.csv()` {#T4_load_csv}

No packages needed. It assumes that your data has headers/column names.

```{r, eval=FALSE}
mydata <- read.csv("filename.csv")
```

If your data has no column names, add `header = FALSE`:

```{r, eval=FALSE}
mydata <- read.csv("filename.csv", header = FALSE)
```

This will name each column V1, V2 etc. You can then rename your columns using the `names()` command.

> **Note:** If you are working with very large files (>100MB or over 100,000 rows), look up the fread() function from the data.table package. It's more complex but MUCH faster.

<br>

#### Excel files (.xlsx, .xls) — `read_excel()` {#T4_load_excel}

Requires the `readxl` package. Install it once via the Packages tab, then add `library(readxl)` to your library code chunk.

```{r, eval=FALSE}
mydata <- read_excel("filename.xlsx")
```

If you want to read in a specific sheet from your Excel file, add the `sheet` argument:

```{r, eval=FALSE}
mydata <- read_excel("filename.xlsx", sheet = "SheetName")
```

<br>

#### TXT files (.txt) — `read.table()` {#T4_load_txt}

No packages needed, but you have to explicitly tell R that the data has headers/column names and that it's "tab separated" (tab character between entries). For weird data-files, can also adjust sep to whatever your actual separater is. 

```{r, eval=FALSE}
mydata <- read.table("filename.txt", header=TRUE, sep="\t")
```

> **Note:** If you are working with very large files (>100MB or over 100,000 rows), look up the fread() function from the data.table package. It's more complex but MUCH faster.

<br>

------------------------------------------------------------------------

### R-specific data {#T4_ReadinRData}

R has its own file formats for saving and reloading R objects. These are useful when you want to save your work mid-analysis and pick it up later, or share processed data with someone else who uses R.

<br>

#### Built-in package datasets — `data()` {#T4_load_builtindata}

Many R packages come with example datasets ready to load. In this case you don't need the file containing the data. Instead add the relevant `library()` to your library code chunk, then use `data()` to load the dataset. This will show up as a "promise". Then run any other command to load it fully. Sometimes you won't see its name in the environment.

```{r, eval=FALSE}
data("penguins")   # load the dataset into your global environment
glimpse(penguins)  # view it
```

> **Important:** Just running `library(somepackage)` makes its datasets *accessible* to R, but they live in the package environment — not your global environment. This means you can accidentally use them without them showing up in your Environment pane or in `ls()`. Always use `data()` explicitly to load a dataset into your global environment, where you can see and work with it properly.

<br>

#### RDS files (.rds) — `readRDS()` {#T4_load_rds}

This is R's own format to store a single variable as a file. RDS saves a **single** R object (a data frame, model output, list, etc.) and reads it back in exactly as you left it. You can assign it any name on import. No packages needed.

```{r, eval=FALSE}
mydata <- readRDS("filename.rds")
```

<br>

#### RData files (.RData, .rda) — `load()` {#T4_load_rdata}

This is R's own format to store many variables as a single file. RData files contain **multiple** R objects at once. Loading it is different from other formats — you use `load()` without `<-`, and the objects reappear in your environment automatically with their original names.

```{r, eval=FALSE}
load("filename.RData")
```

> **Note:** Because the object names are baked into the file, you cannot rename them on import the way you can with `readRDS()`. After loading, check your Environment pane to see what objects have appeared.

<br>

------------------------------------------------------------------------

### Spatial data {#T4_ReadinSpatial}

<br>

#### Vector files (.shp, .geojson, .gpkg) — `st_read()` {#T4_load_vector}

Vector spatial data includes points, lines, and polygons (e.g. country borders, survey locations, river networks). Requires the `sf` package. Install it once via the Packages tab, then add `library(sf)` to your library code chunk.

```{r, eval=FALSE}
mydata <- st_read("filename.shp")      # shapefile
mydata <- st_read("filename.geojson")  # GeoJSON
mydata <- st_read("filename.gpkg")     # GeoPackage
```

`st_read()` works the same way regardless of format — just change the filename and extension.

> **Note for shapefiles:** A shapefile is actually several files sharing the same name (e.g. `.shp`, `.dbf`, `.shx`, `.prj`). All of them need to be in the same folder. You only type the `.shp` filename in your code — R finds the rest automatically.

<br>

#### Raster files (.tif, .nc, .img) — `rast()` {#T4_load_raster}

Raster data is grid-based (e.g. satellite imagery, elevation models, climate surfaces). Requires the `terra` package. Install it once via the Packages tab, then add `library(terra)` to your library code chunk.

```{r, eval=FALSE}
mydata <- rast("filename.tif")   # GeoTIFF (most common)
mydata <- rast("filename.nc")    # NetCDF (common for climate data)
```

If your raster file contains multiple layers (e.g. monthly temperature across 12 months), `rast()` will load them all as a multi-layer object. You can check how many layers you have with `nlyr(mydata)`.

<br>

------------------------------------------------------------------------

## Troubleshooting {#T4_Troubleshooting}

If you get an error when reading in data, work through this checklist before asking for help.

<br>

**Is the filename exactly right?** R is case-sensitive. `Mydata.csv` and `mydata.csv` are different files. Always include the file extension (`.csv`, `.xlsx`, `.shp`, etc.).

<br>

**Is the file in the right place?** The data file needs to be in the **same folder** as your `.Rproj` file, not in Downloads or on your Desktop.

<br>

**Are you running your project?** Check the top-right corner of RStudio — it should show your project name, not "Project: (None)". If it says None, go to File \> Open Project and open your `.Rproj` file.

<br>

**Have you installed and loaded the right package?** `read.csv()`, `readRDS()`, and `load()` need no package. Everything else requires one:

| Function            | Package  |
|---------------------|----------|
| `read_excel()`      | `readxl` |
| `st_read()`         | `sf`     |
| `rast()` / `vect()` | `terra`  |

Installing a package (via the Packages tab) only needs to happen once. Loading it with `library()` needs to happen every session — put your `library()` calls in your setup code chunk at the top of your script.

<br>

**For shapefiles: are ALL the component files present?** A shapefile is made up of multiple files (`.shp`, `.dbf`, `.shx`, and usually `.prj`). If any are missing — for example if you only copied the `.shp` — R will throw an error. Make sure the entire set of files with the same base name are all in your project folder.

<br>

**Your data loaded but looks wrong?** Run `head(mydata)` or `glimpse(mydata)` straight after reading in your data. This helps you catch problems early — for example, all your data ending up in one column is common with CSVs that use `;` instead of `,` as a separator. If that happens, try:

```{r, eval=FALSE}
mydata <- read.csv("filename.csv", sep = ";")
```
