```{r setup, include=FALSE, message=FALSE, warning=FALSE}
# OPTIONS -----------------------------------------------
knitr::opts_chunk$set(echo = TRUE, 
                      warning=FALSE, 
                      message = FALSE)

# Tutorial packages
library(skimr)
library(dplyr)
library(naniar)
library(tidyr)
library(tidyverse)
```

# Missing Data In Depth {#T6B_MissingData}

<br>

This tutorial goes deeper on missing data than the QC steps covered in Tutorial 5. It covers how to identify *why* data is missing, how to visualise it, and how to handle it appropriately depending on your situation — including how missing data interacts with regression models.

> **Before you start:** Make sure you have the following packages installed and loaded in your library code chunk: `skimr`, `dplyr`, `naniar`, `tidyr`, and `tidyverse`.

We will use a small example dataset throughout so you can see exactly what each command does:

```{r}
testdata <- data.frame(
  x = c(1, 2, 3, NA, 5),
  y = c("A", "B", "C", "D", NA),
  z = c(10, 15, NA, 20, 25),
  randomnotes = c(NA, NA, NA, "hello", NA)
)
print(testdata)
```

<br>

---

## What is missing data? {#T6B_WhatIsMissing}

In R, missing values are stored as `NA`. You may also encounter `NaN` (Not a Number), which appears when a calculation produces an undefined result (e.g. dividing zero by zero).

<br>

#### Missing values in external datasets {#T6B_ExternalMissing}

When reading in data from outside R, missing values are not always stored as `NA`. Common placeholders include:

- `-1` (common in government datasets)
- `99`, `-99`, `9999`, `999` (arbitrary placeholders)
- Blank cells in spreadsheets

If you know your dataset uses non-standard codes for missing values, you can tell R what to look for when reading in the file. Always check your data carefully afterwards to make sure it worked correctly.

For CSV files:

```{r, eval=FALSE}
mydata <- read.csv("data.csv", na.strings = c("", "NA", "-1", "99", "9999"))
```

For Excel files:

```{r, eval=FALSE}
mydata <- read_excel("data.xlsx", na = c("", "NA", "-1", "999"))
```

`na.strings` (CSV) and `na` (Excel) both tell R which values to treat as `NA` on import.

<br>

#### Turning existing values into NA {#T6B_ForceNA}

Sometimes you only notice a missing data code after reading the data in. You can overwrite specific values with `NA` directly. For example, to replace any impossible negative prices in a house price dataset:

```{r, eval=FALSE}
HousesNY$Price[HousesNY$Price < 0] <- NA
```

<br>

---

## Why is data missing? {#T6B_WhyMissing}

Understanding *why* data is missing matters — it affects which approach to handling it is appropriate.

**Missing Completely at Random (MCAR)**
The probability of missingness is unrelated to any variable in the dataset. Example: a sensor randomly fails. If data is MCAR, removing missing cases is usually unbiased.

**Missing at Random (MAR)**
Missingness is related to other observed variables, but not to the missing value itself. Example: younger participants are less likely to report their income, but within each age group, missingness is random. Multiple imputation methods can help here.

**Missing Not at Random (MNAR)**
Missingness depends on the value of the missing data itself. Example: people with higher incomes are less likely to report their income. This is the most problematic — the missingness mechanism itself needs to be modelled.

> **Key takeaway:** Only remove missing data casually when you have good reason to believe it is MCAR. For MAR or MNAR, seek specialist statistical advice.

<br>

---

## Identifying missing data {#T6B_Identifying}

<br>

#### Counting NAs {#T6B_Counting}

To count missing values per column:

```{r}
colSums(is.na(testdata))
```

To count missing values per row:

```{r}
rowSums(is.na(testdata))
```

To count NAs in a single column:

```{r}
sum(is.na(testdata$x))
```

To include missing values in a frequency table:

```{r}
table(testdata$y, useNA = "ifany")
```

<br>

#### Summary functions {#T6B_SummaryFunctions}

`summary()` and `skim()` both report missing value counts automatically:

```{r}
summary(testdata)
```

```{r}
skim(testdata)
```

<br>

#### Visualising missing data — `naniar` {#T6B_Visualising}

The `naniar` package provides helpful visualisations for understanding the pattern of missing data across your dataset.

```{r}
gg_miss_var(testdata)   # count of missing values per variable
vis_miss(testdata)      # heatmap showing where NAs are across the whole dataset
```

`vis_miss()` is particularly useful for spotting whether missingness clusters in particular rows or columns — which can give clues about whether data is MCAR, MAR, or MNAR.

<br>

---

## Handling missing data {#T6B_Handling}

<br>

### Remove rows with any NA — `na.omit()` {#T6B_naomit}

**Listwise deletion** removes any row that contains an NA in *any* column:

```{r}
data_complete <- na.omit(testdata)
data_complete
```

> **Warning:** If your dataset has columns with lots of NAs that are irrelevant to your analysis (like a free-text notes column), `na.omit()` will still remove rows because of them. Always check how many rows you lose:
> ```{r, eval=FALSE}
> nrow(testdata)       # before
> nrow(data_complete)  # after
> ```
> If you've lost more than expected, use the targeted approach below instead.

<br>

### Remove rows where specific columns are NA {#T6B_FilterNA}

To remove rows only where particular columns are missing, use `filter()`:

```{r}
data_filtered <- testdata |>
  filter(!is.na(x) & !is.na(z))

data_filtered
```

`!is.na(x)` means *"keep rows where x is NOT missing"*. This is more precise than `na.omit()` and is usually the better choice.

You can also use `complete.cases()` for the same result:

```{r}
data_filtered <- testdata[complete.cases(testdata$x), ]
```

Or `drop_na()` from `tidyr`:

```{r}
data_filtered <- testdata |>
  drop_na(x, z)
```

<br>

### Replace NAs with a value {#T6B_Replace}

Sometimes you want to keep the row but fill in a specific value — for example, replacing `NA` with `0` or `"Unknown"`.

Using `ifelse()`:

```{r}
testdata$x <- ifelse(is.na(testdata$x), 99, testdata$x)
testdata
```

Using `replace_na()` from `tidyr` (cleaner for multiple columns at once):

```{r}
testdata <- testdata |>
  replace_na(list(x = 0, y = "Unknown"))
testdata
```

> **Note:** Replacing NAs with a fixed value is a form of imputation. Only do this if you have a principled reason — for example, replacing `NA` counts with `0` when you know the true value is zero, not unknown.

<br>

---

## Missing data in common functions {#T6B_Functions}

You don't always need to remove missing data in advance. Many R functions have a built-in option to ignore NAs.

<br>

#### `na.rm = TRUE` {#T6B_naRM}

By default, most summary functions return `NA` if any values are missing:

```{r}
mean(testdata$z)
```

Add `na.rm = TRUE` to ignore missing values:

```{r}
mean(testdata$z, na.rm = TRUE)
sd(testdata$z,   na.rm = TRUE)
```

`na.rm = TRUE` works in `mean()`, `median()`, `sd()`, `sum()`, `min()`, `max()`, and most other summary functions.

<br>

#### Correlation and missing data {#T6B_Correlation}

The `cor()` function doesn't use `na.rm` — it has its own argument:

```{r}
cor(testdata$x, testdata$z, use = "complete.obs")
```

For a full correlation matrix, **pairwise deletion** uses all available data for each pair of variables, rather than removing entire rows:

```{r, eval=FALSE}
cor(testdata, use = "pairwise.complete.obs")
```

Pairwise deletion retains more data than listwise deletion, but can result in different sample sizes per correlation — worth noting when interpreting results.

<br>

---

## Missing data in regression models {#T6B_Regression}

Most modelling functions in R remove rows with missing values automatically. The question is *how* they do it — and this matters when you look at predictions.

```{r}
model_omit    <- lm(z ~ x, data = testdata, na.action = na.omit)
model_exclude <- lm(z ~ x, data = testdata, na.action = na.exclude)
```

| Method | What it does | When to use |
|---|---|---|
| `na.omit` | Removes rows with missing data; drops them from predictions | When you don't need predictions for missing rows |
| `na.exclude` | Removes rows with missing data; keeps `NA` in predictions at the original row positions | When you need predictions to align with your original dataset |

**I recommend `na.exclude` unless you have a specific reason not to.**

The difference becomes clear when you look at predictions:

```{r}
predict(model_omit)     # predictions only for complete cases — shorter vector
predict(model_exclude)  # predictions with NA preserved at missing rows — same length as original data
```

`na.exclude` keeps the structure of your dataset intact, which matters in time series or longitudinal analyses where the position of missing values is meaningful.

<br>

---

## Command reference {#T6B_Commands}

A quick-copy summary of all commands covered in this tutorial. Replace `testdata` with your dataset name.

```{r, collapse=TRUE, eval=FALSE}
# -------------------------------------------
# Reading in data with non-standard NA codes
# -------------------------------------------
mydata <- read.csv("data.csv", na.strings = c("", "NA", "-1", "99", "9999"))
mydata <- read_excel("data.xlsx", na = c("", "NA", "-1", "999"))

# Force specific values to NA after reading in
mydata$column[mydata$column < 0] <- NA

# -------------------------------------------
# Identifying missing data
# -------------------------------------------
colSums(is.na(testdata))          # NAs per column
rowSums(is.na(testdata))          # NAs per row
sum(is.na(testdata$x))            # NAs in one column
table(testdata$y, useNA = "ifany") # frequency table including NAs
summary(testdata)                  # summary with NA counts
skim(testdata)                     # detailed summary with NA counts
gg_miss_var(testdata)              # naniar: NAs per variable (plot)
vis_miss(testdata)                 # naniar: NA heatmap (plot)

# -------------------------------------------
# Removing rows with missing data
# -------------------------------------------
data_complete <- na.omit(testdata)                          # remove any row with any NA
data_filtered <- testdata |> filter(!is.na(x) & !is.na(z)) # targeted: specific columns only
data_filtered <- testdata |> drop_na(x, z)                  # tidyr equivalent

# -------------------------------------------
# Replacing NAs with a value
# -------------------------------------------
testdata$x <- ifelse(is.na(testdata$x), 99, testdata$x)         # replace with 99
testdata <- testdata |> replace_na(list(x = 0, y = "Unknown"))   # replace multiple columns

# -------------------------------------------
# Handling NAs inside functions
# -------------------------------------------
mean(testdata$x, na.rm = TRUE)
sd(testdata$x,   na.rm = TRUE)
cor(testdata$x, testdata$z, use = "complete.obs")
cor(testdata, use = "pairwise.complete.obs")

# -------------------------------------------
# Regression with missing data
# -------------------------------------------
lm(z ~ x, data = testdata, na.action = na.omit)    # drops missing rows from predictions
lm(z ~ x, data = testdata, na.action = na.exclude) # preserves NA positions in predictions (recommended)
```

<br>

---

## Troubleshooting {#T6B_Troubleshooting}

<br>

**Function returned `NA` instead of a number**
Your column contains missing values. Add `na.rm = TRUE`:
```{r, eval=FALSE}
mean(testdata$x, na.rm = TRUE)
```

<br>

**`na.omit()` removed far more rows than expected**
You probably have a column with many NAs that isn't relevant to your analysis (e.g. a notes column). Use `filter(!is.na(columnname))` or `drop_na(col1, col2)` to target only the columns that matter.

<br>

**`cor()` returned `NA` even with `na.rm = TRUE`**
`cor()` doesn't use `na.rm`. Use `use = "complete.obs"` or `use = "pairwise.complete.obs"` instead.

<br>

**Predictions from `lm()` are a different length to my dataset**
You used `na.action = na.omit` (the default). Switch to `na.action = na.exclude` to preserve NA positions in the prediction vector so it matches your original data length.

<br>

**`vis_miss()` or `gg_miss_var()` not found**
The `naniar` package isn't loaded. Add `library(naniar)` to your library code chunk.
