```{r setup, include=FALSE,message=FALSE,warning=FALSE}
# OPTIONS -----------------------------------------------
knitr::opts_chunk$set(echo = TRUE, 
                      warning=FALSE, 
                      message = FALSE)
graphics.off()

library(tidyverse)
library(sf)
library(terra)
library(tmap)
library(spatstat)

options(tigris_use_cache = TRUE)

```

# Making maps

## LIBRARIES NEEDED

You need the tmap and basemaps libraries in your code chunk.

## Example Data

In this example, I'm making a case study of Texas redistricting data and using that to make some tmaps.

First lets see if my data file has layers using `st_layers()`.

Note, I seem to have a computer address in my commands. "./index_data/TmapData/.." The dot means "start with the project folder", then it will look inside the subfolders I tell it to until it finds the data.

```{r}
# I use the sf package for this  
st_layers("./index_data/TmapData/TexasRedistricting.gdb")
```

It does! We can use st_read() to read in each layer and save it as a different variable.

```{r}
# Yep! (see the Environment tab)
districts108_2002    <-   st_read("./index_data/TmapData/TexasRedistricting.gdb", layer="districts108_2002")
newDistricts2003     <-   st_read("./index_data/TmapData/TexasRedistricting.gdb", layer="newDistricts2003")
tx_counties          <-   st_read("./index_data/TmapData/TexasRedistricting.gdb", layer="tx_counties")
tx_largerCities      <-   st_read("./index_data/TmapData/TexasRedistricting.gdb", layer="tx_largerCities")
tx_voting108         <-   st_read("./index_data/TmapData/TexasRedistricting.gdb", layer="tx_voting108")
tx_votingcenters     <-   st_read("./index_data/TmapData/TexasRedistricting.gdb", layer="tx_voting108_centers")
mycatlookslikeabat     <-   st_read("./index_data/TmapData/TexasRedistricting.gdb", layer="tx_voting108_centers")
```

Now let's check it worked! I can now refer to that dataset by the name I gave it.

```{r}
# Check it worked! I can now refer to that dataset by the name I gave it
summary(districts108_2002)
```

It's also up to you to choose sensible variable names. This works, but I have no idea what it is

```{r}
head(mycatlookslikeabat)
```

<br><br>

## Quick thematic map (qtm)

I use this for a first quick look but its has many more options.

- Type ?qtm into the console to bring up the help file
- Or give your command to chatGPT/claude and as it to change the color etc.

STOP IF IT STARTS GETTING COMPLICATED! Thats what the main tmap code is for.

```{r}
# This makes a quick map. I normally use this to check that the data isn't messed up
qtm(districts108_2002)
```

```{r}
# and you can add a column fill, by adding any column name with " "
tmap_mode("view")
qtm(districts108_2002,"RepMaj")
```

```{r}
# It will now stay on interactive mode until you turn it off.
# tmap_mode plot turns it back to static mode
tmap_mode("plot")
qtm(districts108_2002, "DISTRICT")
```

<br><br>

## Basic Tmap

Once you get past quick looks, ditch qtm. Instead, use the formal tmap structure

- Load a variable name with tm_shape(),
- Then decide how to plot it (e.g. borders, points, lines, polygons etc.
- Connect each line with +

IMPORTANT - this means that tmap commands are long on several lines! Highlight everything from the first tm_shape until you reach the end of the + then ctrl-enter/cmd-enter to run. Or run the entire code chunk. If you only highlight part of the command, you will get a + in the console. This means that R knows there is more command to come and its waiting for you to add it (esc gets out of this)

```{r}

tm_shape(districts108_2002)  +  # Choose the variable districts108_2002
  tm_polygons(fill="RepMaj") +  # Plot as polygons filled with a color scale from the RepMaj column
tm_shape(tx_largerCities)    +  # Choose the variable tx_largerCities
  tm_dots(size=.5)           +  # Add the cities as dots
tm_check_fix()                  

# Useful command to add at the end, it refreshes the gis data before making the plot

```

<br><br>

## Start adding tmap options

You need the tmap and basemaps libraries in your top code chunk (and have run it)

You can do as much with tmap as you can with ESRI/ARC/GIS software.

For example, here is an "ugly map", with a load of options added so you can identify them. I've written this code twice! Once with just the command and the other with many comments to explain what I did.

```{r}
# This is all one command!  Look for the + symbol

# add an ESRI basemap
tm_basemap() +                
  
# POLYGONS (Filled shapes)
# Districts   
tm_shape(districts108_2002) +  
   tm_polygons(fill="RepMaj",  
               col = "grey", 
               fill.scale = tm_scale_intervals(values = "viridis",  
                                               style = "jenks"),
               fill.alpha = .5,
               fill.legend = tm_legend(title = "Hello!") ) +
   
  
# POINTS (symbols)
# Add cities as symbols
tm_shape(tx_largerCities) +   
  tm_symbols(fill = "#f0daed",
             col="magenta", 
             shape=23, 
             fill.alpha=.5,
             col.legend = tm_legend(title = "Cities in Texas!") ) +
   
   
# POINTS (dots)
# Add voting centres as transparent dots filled with republican vote numbers
tm_shape(tx_votingcenters) +   
  tm_dots(fill = "REP", 
          size=.4,
          fill.alpha=.8,
          fill.scale = tm_scale_intervals(values = "Spectral",  
                                          style = "quantile"),
          fill.legend = tm_legend(title = "Republican votes") ) +
   
  
# POLYGONS (borders)
# add new voting districts as red borders (e.g. no fill)
tm_shape(newDistricts2003) +
  tm_borders(col = "red") +
   
    
# POLYGONS (borders)
# add new voting districts as red borders (e.g. no fill)
tm_shape(newDistricts2003) +
  tm_borders(col = "red") +

# EXTRAS   
# move legend
tm_layout(legend.outside = TRUE, legend.outside.position = "right")+

# Scalebar
tm_scalebar(position = c("left", "bottom")) +
   
# Add compass
tm_compass(position = c("right", "bottom"), type = "arrow") +
   
# Add credits
tm_credits("You can write anything here", 
             position = c("left", "bottom"), 
             alpha = .8, size = 0.6) 

```

<br><br>

## Nicer example

Below this section, I put a load of tmap components so you can customise your own, although I would often start by asking chatGPT (ASK FOR TMAP VERSION 4 CODE) to help

```{r}


#==========================================================================
# [4] MY VERSION (after programming in R for 20+ years) 
# I also asked ChatGPT for the tmap VERSION 4 code to add in any feature I wanted 
# Then asked it "is there a simpler way to write that code" - there often is
#
# because I play with fonts, it will look weird if your plot window is small.
# I normally make the window larger by draging around the central dividers let it resize, and re-run when i am happy
# If you get a label.style error, quit Rstudio, reopen and REINSTALL TMAP 
# (your version is out of date - this is a new feature as of 3 weeks ago)

# To see more basemap codes, see here for the full list and copy their name into the command
# https://leaflet-extras.github.io/leaflet-providers/preview/ 
#==========================================================================

# Add basemap
tm_basemap("CartoDB.Positron") +
   
   # Voter gradient polygons with custom breaks and legend
   tm_shape(districts108_2002, unit = "km") +
   tm_polygons(fill = "RepMaj", 
               lwd = .5, fill_alpha = .75, col_alpha = .8,
               fill.scale = tm_scale_intervals(values = "-brewer.rd_bu",
                                               midpoint = 0,
                                               style = "fixed",
                                               breaks = c(-150000,-100000,
                                                          -50000, 0, 50000,
                                                          100000,150000),
                                               label.style = "continuous"),
               fill.legend = tm_legend(title = "Republican Majority \nin the 2002 \nCongressional \nelections",
                                       group_id = "main_legend",
                                       frame = FALSE,
                                       reverse = TRUE,
                                       text.size = 0.8,
                                       title.align = "left",
                                       title.size = 1.15,
                                       col = "black",
                                       lwd = 1.5,
                                       ticks.lwd = 0,
                                       title.padding = c(.25, 0, 1, 0))) +
   
   # City dots with a slightly lower custom legend the "white" dots are invisible to make it move lower on the page.
   tm_shape(st_geometry(tx_largerCities), unit = "km") +
   tm_dots(size = .18, fill_alpha = .5, col = grey(0.8)) +
   tm_add_legend(type = "symbols",
                 group_id = "main_legend",
                 fill = c("white", "black"), col = c("white", "black"),
                 shape = c(21, 21), size = c(0.2, .25),
                 labels = c(" ", "Urban Centre"), title = "",
                 frame = FALSE, text.size = .9) +
   
   # Add standard map components
   tm_layout(legend.outside = TRUE, legend.outside.position = "right")+
   tm_scalebar(position = c("left", "bottom")) +
   tm_compass(position = c("right", "bottom"), type = "arrow") +
   tm_credits("Basemap: CartoDB Positron. © OpenStreetMap contributors © CARTO", 
              position = c("left", "bottom"), 
              alpha = .8, size = 0.6) +
   tm_credits("Cartography: Dr. Helen Greatrex, Penn State, 2026",
              position = c("left", "bottom"), 
              alpha = .8, size = 0.6) +
   tm_credits("Data: Provided as class example",
              position = c("left", "bottom"), 
              alpha = .8, size = 0.6)



```

<br><br>

## Tmap components

### tmap_mode()

To switch between a static map and an interactive, zoomable map, run `tmap_mode("plot")` or `tmap_mode("view")` BEFORE your map code.

THIS DOES NOT NEED A PLUS SIGN. ITS A SEPARATE COMMAND TO THE COMPUTER

<br>

### Colour palettes

There are many colour palettes available. You choose one using `values =` inside a scale function (see the layer sections below).

- Examples of palette names are `"viridis"`, `"Spectral"` and `"brewer.set2"`.
- To see all the palettes,
  - run `cols4all::c4a_gui()`, which opens an interactive viewer,
  - or `cols4all::c4a_palettes()`, which prints the names.
- To reverse a palette, put a minus sign in front of the name, e.g. `values = "-viridis"`.
- To use your own colours, give a vector using the c command, e.g. `values = c("white", "darkred")`.

<br>

### The + sign

Every layer is connected with a `+` sign. Each bullet below is a "lego brick". Copy the code into your map and change the parts in CAPITALS to match your own data.\

<br>

### tm_basemap()

This adds a background map, such as streets or satellite imagery, underneath your data. It normally goes first.

- To add the default basemap (a light grey map), use `tm_basemap()`.
- There are many basemaps available. To choose one, put its name in quotes, e.g.
  - `tm_basemap("OpenStreetMap")`
  - `tm_basemap("Esri.WorldImagery")` for satellite imagery, or `tm_basemap("Esri.WorldTopoMap")` for a topographic map.
- To see all the basemap options,
  - run `names(leaflet::providers)` in the console,
  - or browse them visually at <https://leaflet-extras.github.io/leaflet-providers/preview/>
- To make the basemap fainter, add alpha (0 = invisible, 1 = solid), e.g.
  - `tm_basemap("OpenStreetMap", alpha = 0.5)`.
  - 

<br>

### IMPORTANT tm_shape()

This MUST come before each layer you want to plot. It is how you load that dataset to map.

- To load a layer, use `tm_shape(YOUR_DATA) +`

  - Then the next command says how you want to plot it\
    <br>

- To zoom the map to the extent of a different layer, use `tm_shape(YOUR_DATA, bbox = OTHER_LAYER)`.

- By default the first `tm_shape()` sets the map extent and projection. To choose a different layer for this, use `tm_shape(YOUR_DATA, is.main = TRUE)`.

<br>

### tm_dots()

These are small, simple points.

- To change the size, use `size`. E.g. `tm_dots(size = 0.5)`.
  - To size the dots by a numeric column, use `tm_dots(size = "NUMERIC_COLUMN")`.
  - This can be annoying and can range from 0.04 to 2 depending on how you want the final map to look

- To fill with a single colour, use `fill` e.g.

  - `tm_dots(fill = "blue")`.

  \

- To fill using colors from a categorical column, use this

  - `tm_dots(fill = "COLUMN_NAME", fill.scale = tm_scale_intervals(values = "brewer.set2", style="jenks"))`.
    - Where values is your color palette (see above)
    - and style is one of `"cat"`, `"sd"`, `"pretty"`, `"kmeans"`, `"hclust"`, `"bclust"`, `"fisher"`, `"jenks"`, `"dpih"` and `"headtails"` \

<!-- -->

- To fill using a numeric column split into classes, use
  - `tm_dots(fill = "NUMERIC_COLUMN", fill.scale = tm_scale_intervals(values = "Spectral", style = "jenks", n = 5))`.
  - Other options for `style` include `"quantile"`, `"equal"`, `"pretty"`, `"log10_pretty"` and `"fixed"`.\

<!-- -->

- OR to fill using colors from a continuous column, use this

  - `tm_dots(fill = "COLUMN_NAME", fill.scale = tm_scale_continuous(values = "brewer.blues"))`.
    - Where values is your color palette (see above)

    - There are also many other options. You can scale both size and color e.g.

      ```         
      tm_scale_continuous(
            midpoint = 1000,
            limits   = c(0, 12e6), 
            ticks    = c(1e5, 3e5, 8e5, 4e6), 
            labels   = c("0-200K", "200-500K", "500-1000K", "1mil or more"), 
            outliers.trunc = c(TRUE, TRUE)))
      ```

- To add transparency (0 = invisible, 1 = solid), use `fill_alpha` e.g. `fill_alpha = 0.5`

- To give the legend a title, add

  - `fill.legend = tm_legend(title = "My title")`.
  - For size legends, use `size.legend = tm_legend(title = "My title")`.

- To hide the legend, add `fill.legend = tm_legend_hide()`.

### tm_symbols()

These are points with a shape, a fill colour and a border colour.

- To set a single fill and border colour, use `tm_symbols(fill = "pink", col = "magenta")`.

- SHAPE: There are many shapes available, set by number, e.g. `tm_symbols(shape = 23)`. Shapes 21 to 25 have a separate fill and border colour. To see the shape numbers, run `?pch` in the console.

- SIZE: To change the size, use `size`. E.g. `tm_dots(size = 0.5)`.

  - To size the dots by a numeric column, use `tm_dots(size = "NUMERIC_COLUMN")`.
  - This can be annoying and can range from 0.04 to 2 depending on how you want the final map to look

- To fill the symbols using a column, use the same `fill` and `fill.scale` code as in `tm_dots()`.

- To change the border thickness, use `tm_symbols(lwd = 2)`. where you change the number

- To add transparency, use `tm_symbols(fill_alpha = 0.5, col_alpha = 0.8)`.

- To give the legend a title, use `fill.legend`, `size.legend` or `shape.legend`, depending on what you mapped to a column, e.g. `shape.legend = tm_legend(title = "Type")`.

<br>

### tm_borders()

These are polygon outlines with no fill.

- To set a single colour, use `tm_borders(col = "red")`.
- To change the line thickness, use `tm_borders(col = "red", lwd = 2)`.
- To make dashed or dotted lines, use `tm_borders(col = "red", lty = "dashed")` or `lty = "dotted"`.
- To add transparency, use `tm_borders(col = "red", col_alpha = 0.5)`.
- To colour the borders using a column, use `tm_borders(col = "CATEGORY_COLUMN", col.legend = tm_legend(title = "My title"))`.

<br>

### tm_polygons()

These are filled shapes with borders.

- To fill with a single colour, use `tm_polygons(fill = "lightblue")`.
- To fill using a categorical column, use `tm_polygons(fill = "CATEGORY_COLUMN", fill.scale = tm_scale_categorical(values = "brewer.set2"))`.
- To fill using a numeric column with a smooth colour ramp, use `tm_polygons(fill = "RepMaj", fill.scale = tm_scale_continuous(values = "viridis"))`.
- To fill using a numeric column split into classes, use `tm_polygons(fill = "RepMaj", fill.scale = tm_scale_intervals(values = "viridis", style = "jenks", n = 5))`.
- To choose your own class breaks, use `fill.scale = tm_scale_intervals(breaks = c(0, 25, 50, 75, 100), values = "viridis")`.
- To change the border colour and width, use `tm_polygons(fill = "RepMaj", col = "grey", lwd = 0.5)`.
- To remove the borders, use `tm_polygons(fill = "RepMaj", col = NA)`.
- To add transparency, use `tm_polygons(fill = "RepMaj", fill_alpha = 0.5)`.
- To give the legend a title, add `fill.legend = tm_legend(title = "My title")`.
- To hide the legend, add `fill.legend = tm_legend_hide()`.

A legend only appears when something is linked to a column, e.g. `fill = "RepMaj"`. A single colour such as `fill = "blue"` gets no legend.

<br>

### Extras

- To add a title, use `tm_title("My map title")`.
- To put all the legends outside the map on the right, add `tm_place_legends_right()`. There are also `tm_place_legends_left()`, `tm_place_legends_top()`, `tm_place_legends_bottom()` and `tm_place_legends_inside()`.
- To move one legend inside the map, add `fill.legend = tm_legend(position = tm_pos_in("left", "top"))` to that layer.
- To remove the frame around the map, use `tm_layout(frame = FALSE)`.
- To change the background colour, use `tm_layout(bg.color = "lightgrey")`.
- To add a scalebar, use `tm_scalebar(position = c("left", "bottom"))`. To set the distances shown, add e.g. `breaks = c(0, 100, 200)`.
- To add a compass, use `tm_compass(type = "arrow", position = c("right", "bottom"))`. Other types are `"4star"`, `"8star"`, `"radar"` and `"rose"`.
- To add credits or a data source, use `tm_credits("Data: US Census", position = c("left", "bottom"), size = 0.6)`.
- To add text labels from a column, use `tm_shape(YOUR_DATA) + tm_text("NAME_COLUMN", size = 0.6)`.
- To save a map, first save it as a variable, e.g. `mymap <- tm_shape(YOUR_DATA) + tm_polygons()`, then run `tmap_save(mymap, "mymap.png")`.

## 
